Fast Bayesian Inference for Exoplanet Discovery in Radial Velocity Data

### 2014-12-31 Paper/ms.tex
mention orbital stability
Prior table names arguments

An apparently strange choice is the conditional prior for the (logarithms of)
the orbital
periods, which is a biexponential distribution given a location parameter
$\mu_P$ and a scale parameter $w_P$ which determines the width of the
distribution\footnote{A biexponential distribution with location parameter
$\mu$ and scale parameter $w$ has probability density function
$p(x|\mu, w) = \frac{1}{2w} \exp\left(-\frac{|x - \mu|}{w}\right)$.}.
Rather than assigning independent priors to the log periods, the hierarchical
model allows for the periods to ``cluster around'' a typical period $\mu_P$ if
there is evidence for this. On the other hand, independent priors for the
periods would imply a strong prior commitment to the hypothesis that the
periods are spread out across the whole prior volume (equivalent to assuming
a fixed large value for $w_P$).
A more conventional choice for the conditional prior given $\mu_P$ and $w_P$
would have been a normal distribution. However, the \citet{rjobject} software
needs to know the corresponding cumulative distribution and its inverse, which
are not available in closed form for the normal distribution.

For the velocity semi-amplitudes $\{A_i\}$, we chose an exponential distribution
given the hyperparameter $\mu_A$. Our prior for $\mu_A$ spans many orders of
magnitude but expresses a slight preference for $\mu_A$ being of order unity,
using a Cauchy distribution. The prior for the semi-amplitudes
will influence how many of these low amplitude planets will be inferred: if we
believe there are many, and the data are uninformative about low amplitude
planets, then the posterior distribution for $N$ will also indicate that there
may be many low amplitude planets. However, their other properties, such as
their orbital periods, will not be well determined. The Beta prior for
eccentricity was suggested by Gregory (priv. comm) and is an approximation
to the inferred frequency distribution of eccentricities in the population
\citep{kipping}.

\begin{table*}
\begin{tabular}{|l|l|l|}
\hline
Quantity	&	Meaning		& Prior\\
\hline
{\bf Hyperparameters}	&	\\
$N$		& Number of planets	& Uniform$(\{0, 1, ..., N_{\rm max}\})$\\
$\mu_P$		&	Median orbital period	& $\log(\mu_P) \sim$ Cauchy$(5.9, 1)T(-15.3, 27.1)$\\
$w_P$		&	Diversity of orbital periods & $w_P \sim$ Uniform$(0.1, 3)$\\
$\mu_A$		&	Mean amplitude	& $\log(\mu_A) \sim$ Cauchy$(0, 1)T(-21.2, 21.2)$\\
\hline
{\bf Planet Parameters}\\
$P_i$		&	Orbital period	&	$\log(P_i) \sim $ Biexponential($\log(\mu_P), w_P$)\\
$A_i$		&	Semi-mplitude of signal	& Exponential$(\textnormal{mean}=\mu_A)$\\
$\phi_i$	&	Phase of signal	&	Uniform($0, 2\pi$)\\
$e_i$		&	Eccentricity of orbit	&	Beta$(1, 3.1)T(0, 0.8)$\\
$\omega_i$	&	Viewing angle	&	Uniform$(0, 2\pi)$\\
\hline
{\bf Other}\\
$m_0$		&	Constant DC offset	&	Uniform$(m_{0, \rm min}, m_{0, \rm max})$\\
$\sigma_{\rm extra}$	& ``Extra noise'' parameter	& $\log(\sigma) \sim$ Cauchy$(0, 1)T(-21.2, 21.2)$\\
$\nu$		& Shape parameter for $t$-distribution for noise & $\log(\nu) \sim$ Uniform$(\log(0.01), \log(1000))$\\
\hline
{\bf Data}\\
$Y_i$		& Radial velocity measurements	&
		Student-$t\left(m(t_i), \sqrt{\sigma_i^2 + \sigma_{\rm extra}^2}, \nu\right)$
\end{tabular}
\caption{All of the prior distributions in our Bayesian model.
The priors for the planet parameters are defined conditional on the values
of the hyperparameters. Uniform priors were used for parameters like phases.
For parameters where a rough initial guess is possible, heavy-tailed Cauchy
distributions were used so this information could be taken into account
in a non-dogmatic way. Time units are in days and amplitude units are
in metres per second. The maximum number of planets, $N_{\rm max}$, was set
to 10. The prior limits for $m_0$ were set to the minimum and maximum value
in the dataset, which is not strictly a valid strategy. The 
Truncation intervals are specified with the $T(a,b)$ notation.\label{tab:priors}}
\end{table*}

@@

It is now recognised that
there are many possible sources for oscillations in a data set and not all
such oscillations should be claimed as planets \citep{2014Sci...345..440R}. 
Our model cannot distinguish between oscillations due to planets and
oscillations due to stellar activity: any oscillations found in the dataset
will be described as ``planets'' by the model. However it is interesting that
we find many more signals in the data than previous authors.
By inspecting the posterior distribution for the periods
(Figure~\ref{fig:gliese581_periods}), we see that only four of the periods
are well determined, corresponding to the known periods of Gliese 581 b, c, d,
and e. The other ``periods'' are very uncertain. It is quite likely that a
non-sinusoidal signal due to stellar activity is being modelled as several
periods \citep{astero}, and if the model were extended to include a ``stochastic''
oscillation \citep{gaussproc}, the number of periods detected may be reduced
substantially. An example model fit for Gliese 581 is shown in
Figure~\ref{fig:gliese581}.

@@

>>>>

An apparently strange choice is the conditional prior for the (logarithms of)
the orbital
periods, which is a biexponential distribution given a location parameter
$\mu_P$ and a scale parameter $w_P$ which determines the width of the
distribution\footnote{A biexponential distribution with location parameter
$\mu$ and scale parameter $w$ has probability density function
$p(x|\mu, w) = \frac{1}{2w} \exp\left(-\frac{|x - \mu|}{w}\right)$.}.
Rather than assigning independent priors to the log periods, the hierarchical
model allows for the periods to ``cluster around'' a typical period $\mu_P$ if
there is evidence for this. On the other hand, independent priors for the
periods would imply a strong prior commitment to the hypothesis that the
periods are spread out across the whole prior volume (equivalent to assuming
a fixed large value for $w_P$).
A more conventional choice for the conditional prior given $\mu_P$ and $w_P$
would have been a normal distribution. However, the \citet{rjobject} software
needs to know the corresponding cumulative distribution and its inverse, which
are not available in closed form for the normal distribution.
Our prior for $w_P$, which controls the diversity of the log-periods, was
uniform between 0.1 and 3, since it is unlikely that many planets have
extremely similar or extremely different (over several orders of magnitude)
orbital periods.

For the velocity semi-amplitudes $\{A_i\}$, we chose an exponential distribution
given the hyperparameter $\mu_A$ which sets the mean of the exponential distribution.
Our prior for $\mu_A$ spans many orders of
magnitude but expresses a slight preference for $\mu_A$ being of order unity,
using a Cauchy distribution. The prior for the semi-amplitudes
will influence how many of these low amplitude planets will be inferred: if we
believe there are many, and the data are uninformative about low amplitude
planets, then the posterior distribution for $N$ will also indicate that there
may be many low amplitude planets. However, their other properties, such as
their orbital periods, will not be well determined. The Beta prior for
eccentricity was suggested by Gregory (priv. comm) and is an approximation
to the inferred frequency distribution of eccentricities in the population
\citep{kipping}.

\begin{table*}
\begin{tabular}{|l|l|l|}
\hline
Quantity	&	Meaning		& Prior\\
\hline
{\bf Hyperparameters}	&	\\
$N$		& Number of planets	& Uniform$(\{0, 1, ..., N_{\rm max}\})$\\
$\mu_P$		&	Median orbital period (years)	& $\log(\mu_P) \sim$ Cauchy$({\rm location}=5.9, {\rm scale}=1)T(-15.3, 27.1)$\\
$w_P$		&	Diversity of orbital periods & $w_P \sim$ Uniform$({\rm min}=0.1, {\rm max}=3)$\\
$\mu_A$		&	Mean amplitude (metres per second)	& $\log(\mu_A) \sim$ Cauchy$({\rm location}=0, {\rm scale}=1)T(-21.2, 21.2)$\\
\hline
{\bf Planet Parameters}\\
$P_i$		&	Orbital period	&	$\log(P_i) \sim $ Biexponential$({\rm location}=\log(\mu_P), {\rm scale}=w_P$)\\
$A_i$		&	Semi-amplitude of signal	& Exponential$(\textnormal{mean}=\mu_A)$\\
$\phi_i$	&	Phase of signal	&	Uniform($\textnormal{min}=0, \textnormal{max}=2\pi$)\\
$e_i$		&	Eccentricity of orbit	&	Beta$(\alpha=1, \beta=3.1)T(0, 0.8)$\\
$\omega_i$	&	Viewing angle	&	Uniform$({\rm min}=0, {\rm max}=2\pi)$\\
\hline
{\bf Other}\\
$m_0$		&	Constant DC offset	&	Uniform$({\rm min}=m_{0, \rm min}, {\rm max}=m_{0, \rm max})$\\
$\sigma_{\rm extra}$	& ``Extra noise'' parameter	& $\log(\sigma) \sim$ Cauchy$({\rm location}=0, {\rm scale}=1)T(-21.2, 21.2)$\\
$\nu$		& Shape parameter for $t$-distribution for noise & $\log(\nu) \sim$ Uniform$({\rm min}=\log(0.01), {\rm max}=\log(1000))$\\
\hline
{\bf Data}\\
$Y_i$		& Radial velocity measurements	&
		Student-$t\left({\rm location}=m(t_i), {\rm scale}=\sqrt{\sigma_i^2 + \sigma_{\rm extra}^2}, {\rm shape}=\nu\right)$
\end{tabular}
\caption{All of the prior distributions in our Bayesian model.
The priors for the planet parameters are defined conditional on the values
of the hyperparameters. Uniform priors were used for parameters like phases.
For parameters where a rough initial guess is possible, heavy-tailed Cauchy
distributions were used so this information could be taken into account
in a non-dogmatic way. Time units are in days and amplitude units are
in metres per second. The maximum number of planets, $N_{\rm max}$, was set
to 10. The prior limits for $m_0$ were set to the minimum and maximum value
in the dataset, which is not strictly a valid strategy. Some of the distributions
were truncated for numerical reasons, the truncated intervals are specified with the
$T(a,b)$ notation.\label{tab:priors}}
\end{table*}

@@

It is now recognised that
there are many possible sources for oscillations in a data set and not all
such oscillations should be claimed as planets \citep{2014Sci...345..440R}. 
Our model cannot distinguish between oscillations due to planets and
oscillations due to stellar activity: any oscillations found in the dataset
will be described as ``planets'' by the model. Another important consideration
is the physical stability of the orbital system which is ignored by this type
of analysis \citep{orbital}. However it is interesting that
we find many more signals in the data than previous authors.
By inspecting the posterior distribution for the periods
(Figure~\ref{fig:gliese581_periods}), we see that only four of the periods
are well determined, corresponding to the known periods of Gliese 581 b, c, d,
and e. The other ``periods'' are very uncertain. It is quite likely that a
non-sinusoidal signal due to stellar activity is being modelled as several
periods \citep{astero}, and if the model were extended to include a ``stochastic''
oscillation \citep{gaussproc}, the number of periods detected may be reduced
substantially. An example model fit for Gliese 581 is shown in
Figure~\ref{fig:gliese581}.

@@

\bibitem[\protect\citeauthoryear{T{\'o}th 
\& Nagy}{2014}]{orbital} T{\'o}th Z., Nagy I., 2014, MNRAS, 442, 454
